---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  echo = FALSE
)
```

```{r setup}
# library(aasim)
devtools::load_all()
simTest <- simulatefromexcel("../tests/testthat/data/siminput.xlsx")
# print(getwd())
# list.files("../tests/testthat/data/")
# print(testthat::test_path())
simTest$sim1$description
```
The purpose of this package is to address the question: How long will an investor's money last? 
Classic financial planning tools assume one age at death, usually a conservative one such as 
95 years old.   Such tools answer the question "Will your money last until you are 95".  
Our approach considers many possible ages of death and weights the results according to 
the probability of living to those ages.   Thus our tool can answer the question, 
"What is the probability your money outlasts you?".  In addition, it can answer the
classic question.  In addition, we offer an additional analysis to answer the question,
"Using historical data, how long would my money have lasted?".

A simulation requires inputs about the investor and the market.  Essentially, we need a starting 
investment and subsequent cash flows.   Then we apply a set of returns to those cash flows
to calculate portfolio values.   If the portfolio meets or exceeds a target value at a specified time, it
is counted as a success.   

### Simulation Inputs
A simulation requires some of the following inputs.  Some inputs are not required
depending on the values of other inputs.  There is general information
associated with a simulation; there may be 0, 1 or 2 persons associated with a 
simulation, and there may be 0 or any number of cash flows.

description
: A description of the simulation.    

nTrials
: The number of trials in a simulation. This applies when the *returnGeneratorMethod* is "S" or "H". 
It is ignored if the *returnGeneratorMethod* is "C".    
 
startValue
: Starting value (initial investment).    

lengthType
: Either "F" for fixed length or "M" for mortality.  If "F", then every trial in the 
simulation will be *length* years long. If "M", the length will depend on the 
mortality tables.  If "F", then *returnGeneratorMethod* may be "S", "H", or "C".  If
"M" it may only be "S" or "H".    

length
: The length in years of all the trials when the *lengthType* is "F".  It is
ignored is the *lengthType* is "M".  In the results of a simulation, there will
be *length* + 1 values because the first value represents year 0 (now) and then 
there are 1 to *length* subsequent years.  The results will have *length* rates
of return.    

seed
: The seed used in the set.seed function to initialize the random number 
generator.  Using the same seed allows one to generate the same sequence of 
random numbers allowing for repeatability.     

defaultInflation
: The default inflation rate which is applied to cash flows.   Each cash flow
may have its own inflation rate.    

ror
: The rate of return used when creating lognormally distributed random returns. 
Enter 8% as 0.08.    

stdDev
: The standard deviation used when creating lognormally distributed random returns. 
Enter 9% as 0.09.    
             
targetValue
: The target value defines the success or failure of a trial.  A trial is a success
if the terminal value exceeds the *targetValue*.  The default is .Machine$double.eps
which is a small positive value. See also *targetValueIsReal*.    

targetValueIsReal
: A boolean to indicate whether the *targetValue* is adjusted by inflation.  
The default is TRUE.    

stockWt
: When calculating returns using historical data, this is the weighting to assign
to stocks.  (1 - *stockWt*) will be assigned to bonds.    

nConsecMonths
: When generating random returns using historical data, this variable indicates the
number of consecutive months which will be drawn.  The default is 12.  Valid 
values are 1, 2, 3, 4, 6, 12 (factors of 12).  For example if *nConsecMonths* is
3, then 4 draws of 3 consecutive months will be used to create a one-year return.    

retAdj
: The random returns generated using historical data (*returnGeneratorMethod* = "H") will be
adjusted by adding *retAdj* to each year's return.  The default is zero. 
Ignored if *returnGeneratorMethod* = "S" or "C".

minDate
: Used to filter the historical data set.   *minDate*  represents the earliest
date historical returns will be used. The default is to use the earliest 
available.      

maxDate = max(sbi$Month),
: Used to filter the historical data set.   *maxDate*  represents the latest
date historical returns will be used. The default is to use the most recent
available.    

overrideInflation
: This applies when *returnGeneratorMethod* = "H".  If TRUE, then the historical
inflation rate is used to adjust cash flows. It is ignored for other 
*returnGeneratorMethod* values. Historical inflation is always used when 
Chronological history is used and never when returns are generated
statistically.

asOfDate
: The date used to calculate current ages of persons in the simulation.  
The default is the Sys.Date(),    

returnGeneratorMethod
: This may be "S", "H", or "C".  "S" causes the system to generate lognormally
distributed random returns.    "H" randomly draws *nConsecMonth* of returns and
inflation using historical data.   "C" is not random. It indicated "Chronological"
history.   This is used to see how long the money will last using history as a guide.    

Each simulation may have 0, 1 or 2 persons.   If 0, then the *lengthType* must be
"F".  Without people, the mortality tables cannot be applied so the *length* will
determine the time horizon of a simulation.

**Persons**
Each person has the following information (variables).

name
: Name of the person.    

initials
: Initials to use in tables, charts where the name will not fit.    

birthdate
: Date of birth.  Used to determine age.  To protect PII, this may be close.    

gender
: M for male or F for female.     

retireAge
: Age of retirement

mort.factor
: Mortality factor.  Values (probabilities of dying) in the mortality table are
multiplied by this value. Larger than 1 values increase chance of dying so 
persons don't live as long as those with smaller values.    

mort.adj.years
: A value to add to each randomly generated age at death.  It may be negative to
reduce lifespans. This is probably the easiest way to adust mortality tables.

**Cash Flows**
Each simulation may have multiple cash flows associated with it.  The variables
associated with a cash flow are:

description
: A string describing the cash flow.      

startType
: Here are the types that describe when a cash flow starts- yr
indicates a numeric value will be supplied such as 1 to start year one and 10
to end year 10. p1age, p2age is the age of person1 or person 2; p1ret, p2ret
is the retirement age of person1 or person 2; p1death, p2death is the age at
death  of person1 or person 2; 1stdeath, 2nddeath is the death of the 1st to
die, or the 2nd to die; start is first year (equivalent to startType=='yr' and
specifying 1 for the start); end is last year.    

A note on the startType and endType.  yr, start, and end may be used in any
simulation regardless of the *lengthType* or *returnGeneratorMethod*.  p1age, p2age,
p1ret, p2ret, may be used in any simulation that defines one or two persons 
respectively.   p1death, p2death, 1stdeath, and 2nddeath may only be used in 
simulations when the *lengthType* is "M" and the *returnGeneratorMethod* is not "C" and 
the appropriate number of persons is defined.

start
: The year or age the cash flow starts depending on the value of 
*startType*.  Ignored in some cases such as 1stdeath.    

endType
: Here are the types that describe when a cash flow starts- yr
indicates a numeric value will be supplied such as 1 to start year one and 10
to end year 10. p1age, p2age is the age of person1 or person 2; p1ret, p2ret
is the retirement age of person1 or person 2; p1death, p2death is the age at
death  of person1 or person 2; 1stdeath, 2nddeath is the death of the 1st to
die, or the 2nd to die; start is first year (equivalent to startType=='yr' and
specifying 1 for the start); end is last year.    

end
: The year or age the cash flow ends depending on the value of 
*endType*.  Ignored in some cases such as 1stdeath.    

type
: Either "c" for contribution or "w" for withdrawal.    

amount
: Amount of the cash flow.    

defaultInflationAdj
: Boolean.  If TRUE, the *defaultInflation* value for the simulation is used.

inflation
: If the *defaultInflationAdj* is FALSE, then this value is used to inflate the
cash flows.

Cash flows in a simulation may be validated using the *validateCashFlows*
function which validates all the cash flows in a simulation. 

### Overview of the Calculations.     

First, the number of trials is established.  If *returnGeneratorMethod* is "S" or "H" 
there will be *nTrials*.  If the value is "C", the number of trials will be a 
function of the number of months of historical data available, the *minDate*, 
the *maxDate* and the *length* of a trial.   Let's say there are 360 months of
data available after adjusting for the *minDate* and *maxDate*.  If the *length*
of a trial is 30 years (360 months) then there is only 1 trial available. If 
361 months of data are available, then there would be 2 trials.

If the *lengthType* is "M" (then there must be 1 or 2 persons as inputs), 
*nTrials* ages of death are generated.   

There is a loop for each trial. *length* + 1 portfolio values will be calculated.
The first value will be set equal to *startValue*.  This may also be considered
a cash flow in year 0.

All the cash flows are evaluated and combined for a trial to create a vector of
*length* + 1.  

The returns are generated for the trial using either: randomly generated from a
lognormal distribution, randomly chosen from history, or using chronological 
history.

The returns are applied the the previous portfolio value and then the result is 
adjusted for the net cash flow.  This means that cash flows occur at the end
of the year.

### Intermediate Output
The simulateWealth function (which is a misnomer in when the *returnGeneratorMethod*
is "C") produces the following results.  We're calling this intermediate output
because it may be used to create final output rather than being displayed to the
user. After the simulateWealth function is run, a list named simulation will be
part of the sim object.  It will contain:

lengths
: A vector with length *nTrials* if *randReturnTYpe* is "S" or "H".  If "C", the
length will be determined by the number of months of historical returns 
available less the *length* + 1.      

agesDeath1 and agesDeath2
: If the person inputs are not supplied, these will be NA.  Otherwise, these 
will be vectors of the ages at which the persons dies.     

portfolioValues
: A list with a vector of the portfolio values for each trial.  The length of
each vector will be one longer than the length of the trials in years.  
The first element represents the starting value plus any cash flows that 
happen immediately (e.g., if the person is retired and a flow starts at 
retirement).    

ratesOfReturns
: A list with a vector of 1 + the returns for each trial.  The length of
each vector will equal the length of the trials in years.     

inflationHist
: A list with a vector of 1 + the historical inflation for each trial.  The length of
each vector will equal the length of the trials in years.  This won't be 
populated if the *randReturnTYpe* is "S".    

cashFlows
: A list with a vector of the net cash flows for each trial.  The length of
each vector will be one longer than the length of the trials in years.  
Contributions will be positive, and withdrawals negative.    

runDate
: The date the simulation was run.

### Example: Chronological History
```{r}
sim <- simTest$ChronHist1
print.sim(sim)

```

